SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2012-12-30 paper/paper.tex
paper: outline, refactoring.

\todo{note constants are user-defined arguments, other simplifications}

\section{Introduction}\label{sec:intro}
The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Nowadays, QuickCheck is a standard
aspect of the Haskell development cycle.  Because QuickCheck generates random
values for testing, counterexamples it finds may be substantially larger
than a minimal counterexample.  In their original QuickCheck paper~\cite{qc},
the authors report the following user experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that presents a method
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a]
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values (according to
some measure) from a given counterexample to attempt to find a smaller
counterexample.  Today, \ttp{smaller} is called \ttp{shrink}.

@@

-- Predicates for defining properties
pre :: B -> Bool
pre (B a b c d) = and \$ map pre' [a, b, c, d]
  where pre' x = add x < 16

post :: B -> Bool
post (B a b c d) =
  add a + add b + add c + add d < 64
  where add :: [A] -> Int16
        add = sum . map (\(\backslash\)(A i) -> i)

@@

Thus, it pays to define \ttp{shrink}.  Unfortunately, even for a simple property
and program like this one, the definition for \ttp{shrink} given in
Figure~\ref{fig:initial} produces an intractable number of potential
counterexamples to test; with the definitions and property provided, which also
rely on QuickCheck's default instances for the list type, the list of potential
counterexamples generated by \ttp{shrink} can contain more than $10^{10}$
elements.  This is an enormous state space to search for a smaller example and
can cause the shrinking phase to take orders-of-magnitude more time than the
counterexample discovery phase.

@@

Typically, a user tries a brute-force approach to reduce the number of potential
counterexamples, for example, by truncating lists using the \ttp{take n} function
that returns the first $n$ elements of a list.  The trade-off is quicker
shrinking with a lower probability of finding a smaller counterexample.  For
example, redefining shrink for type \ttp{B} as follows
%

@@

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Table~\ref{table:results}, giving the results for using QuickCheck and
SmartCheck on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck,
since as we noted, it is infeasible for an example like this).  In the first
three columns, the results of using QuickCheck to generate counterexamples are
shown.  The first column contains the results of using QuickCheck without
shrinking to provide a baseline.  The next two columns show the result of using
QuickCheck with definitions of \ttp{shrink} that limit the lists of fields of
\ttp{B} to 10 (Figure~\ref{lst:newshrink}) and 20 elements
(Figure~\ref{lst:newshrink}, with \ttp{len = 20}), respectively.  (Recall that
with the original definition for shrink from Figure~\ref{fig:initial}, some
counterexamples cause QuickCheck to not return after hours of computation.)  In
each row, we show the maximum, mean, and median number of seconds required by
QuickCheck to generate a counterexample over 100 runs.  In the last row, we show
the mean size of the final value returned, measured by the taking the cumulative
length of the lists of the fields of \ttp{B}.  Notice that while typically
counterexamples are returned in a fraction of a second, there are outliers
taking 21 seconds and 125 seconds for the two tests.  Both tests roughly cut the
size of the counterexample in half; increasing the number of potential
counterexamples can result in a significant performance penalty for this example
without much benefit.

@@

\paragraph{Outline.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Shrinking Data}
In this section, we describe how to efficiently shrink data values.  Before we
begin, recall the basic approach implemented by the \ttp{shrink} method:
generate a list of values, each of which is smaller than the current
counterexample.  Each of the new values generated has nothing in common with the
original counterexample, since each one is freshly generated by recursive
size-constrained calls to \ttp{arbitrary}.

Consider the problem of finding a small counterexample as a graph search
problem.  The set of vertexes are all possible well-typed values (the same as
the counterexample).  There is an edge from vertex $v$ to $v'$ if $v'$ is
smaller (by some measure).  The goal is to fine a vertex from the set of
smallest counterexamples in the graph.

The approach taken by \ttp{shrink} is the most naive graph search algorithm:
it simply generates a random set of potential vertexes to check in the graph.
However, it does not exploit that the original counterexample \ttp{cex} is
already in a neighborhood of likely counterexamples.  So rather than choose
random vertexes, it may be more efficient to test other values that are closely
related to \ttp{cex}.  By restricting the search to \ttp{cex}'s neighborhood, a
smaller counterexample might be missed, and only a local minimum is returned.
However, with a large state space, finding a global minimum by random sampling
vertexes in the graph is possible but too inefficient, as we saw from the
example in the introduction.

To present the algorithm, consider the \ttp{SubTypes} type class below for which
QuickCheck's \ttp{Arbitrary} class is a super-class.\footnote{This presentation
  deviates from the implementation; for example, we elide details regarding
  error handling.}
%
\begin{code}
type Idx = Int
type Size = Int

data SubVal =
  forall a. SubTypes a => SubVal a

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
\end{code}
%
\noindent
\ttp{SubTypes} has five methods.  The \ttp{size} method returns the size of a
value, \ttp{index} returns a sub-value at an breadth-first index in a value,
\ttp{replace} replaces a sub-value at a particular focus, \ttp{constr} returns
the constructor tagging the value, and \ttp{constrs} returns a list of unique
identifiers for the constructors associated with each variant of the type.

@@

constr  tree = ["B"]
constrs tree = ["B", "L"]
constrs L    = ["B", "L"]
\end{code}
%
\noindent
We will sometimes refer to the unwrapped value returned by \ttp{index a i} as
the \emph{$i$th sub-value of $a$}, so for example, \ttp{B\sub{4} L\sub{6}
  L\sub{8}} is the 4th sub-value of \ttp{tree}.  An invariant of \ttp{index} is
that for any value \ttp{a}, consider the least $i \geq 0$ such that \ttp{index a
  i == Nothing}.  For all $0 \leq j < i$, \ttp{index a j /= Nothing}.  We use
this invariant as a termination case in recursive algorithms over the sub-values
of a value.

@@

@@

\section{Counterexample Generalization} \label{sec:generalization}

@@

\subsection{Universal Sub-Value Generalization}
An example best explains the first kind of quantified formulas we generate.
Compare the following formula to the counterexample in Figure~\ref{fig:div-cex}:
%
\begin{code}
\ttp{forall x, y . Div (Div x (Div (C 2) (C 13))) y}
\end{code}
%
\noindent
Here, the formula claims that for all well-typed values \ttp{x} and \ttp{y}, the
property fails.  Not only does the generalization assist the programmer's
insight, but it reduces the sheer size of the counterexample, and it can do so
significantly.  We call the kind of formula just shown \emph{universal sub-value
  generalization} and is implemented in SmartCheck.

@@

extrapolate :: SubTypes a
            => a -> (a -> Property) -> IO [Idx]
extrapolate cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs
    | subTrees cex idx idxs
    = extrapolate' (idx+1) idxs
  extrapolate' idx idxs =
    case index cex idx of
      Nothing -> return idxs
      Just v  -> do
        vs <- newVals v scMaxSize scMaxExtrap
        extrapolate' (idx+1)
          (if allFail cex idx vs prop
             then idx:idxs else idxs)

@@

sumTest :: SubTypes a
          => a -> (a -> Property) -> [Idx] -> IO [Idx]
sumTest cex prop exIdxs = sumTest' 1 []
  where
  sumTest' idx idxs
    | subTrees cex idx (exIdxs ++ idxs)
    = sumTest' (idx+1) idxs
  sumTest' idx idxs =
    case index cex idx of
      Nothing -> return idxs
      Just v  -> do
        vs <- newVals v scMaxSize scConstrMax
        sumTest' (idx+1)
          (if constrFail cex idx vs prop
                (subConstr v) (subConstrs v)
             then idx:idxs else idxs)

constrFail :: SubTypes a => a -> Idx -> [SubVal]
  -> (a -> Property) -> String -> [String] -> Bool
constrFail cex idx vs prop con allCons =
  constrFail' [con] vs
  where
  constrFail' cons vs'
    | length cons == length allCons
    = True
    | null vs'
    = False
    | go v == Just True
    = constrFail' cs (tail vs')
    | otherwise
    = constrFail' cons (tail vs')

    where
    v  = head vs'
    c' = subConstr v
    cs = if c' `elem` allCons
           then cons else c':cons

  go = fail prop . replace cex idx
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:constrs}
\end{figure}

The algorithm

@@

\section{Experiments}

@@

%-------------------------------------------------------------------------------
\section{Related Work}
\label{sec:related}

@@

\cite{}

%-------------------------------------------------------------------------------
\section{Conclusions}
\label{sec:conclusions}

>>>>

\todo{define variant, tag}
\todo{note constants are user-defined arguments, other simplifications}

\section{Introduction}\label{sec:intro}
The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Originally designed for Haskell,
QuickCheck has been ported to other languages and is a now a widely-used testing
tool.  Because QuickCheck generates random values for testing, counterexamples
it finds may be substantially larger than a minimal counterexample.  In their
original QuickCheck paper~\cite{qc}, the authors report the following user
experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that presents a method
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a]
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values (according to
some measure) from a given counterexample to attempt to find a smaller
counterexample.  Today, \ttp{smaller} is called \ttp{shrink}.

@@

-- Predicates for defining properties
add :: [A] -> Int16
add = sum . map (\(\backslash\)(A i) -> i)

pre :: B -> Bool
pre (B a b c d) =
  all (\(\backslash\) -> add x < 16) [a, b, c, d]

post :: B -> Bool
post (B a b c d) =
  add a + add b + add c + add d < 64

@@

Thus, it pays to define \ttp{shrink}.  Unfortunately, even for a simple property
and program like this one, the definition for \ttp{shrink} given in
Figure~\ref{fig:initial} produces an intractable number of potential
counterexamples to test; with the definitions and property provided (together
with QuickCheck's default \ttp{Arbitrary} instance for lists), the list of
potential counterexamples generated by \ttp{shrink} can contain more than
$10^{10}$ elements.  This is an enormous state space to search for a smaller
example and can cause the shrinking phase to take orders-of-magnitude more time
than the counterexample discovery phase.

@@

Typically, a user tries a brute-force approach to reduce the number of potential
counterexamples, for example, by truncating lists using the (\ttp{take n}) function
that returns the first $n$ elements of a list.  The trade-off is quicker
shrinking with a lower probability of finding a smaller counterexample.  For
example, redefining shrink for type \ttp{B} as follows
%

@@

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Table~\ref{table:results}, giving the results for using QuickCheck and
SmartCheck on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck,
since as we noted, it is not even feasible for an example like this).  In the
first three columns, the results of using QuickCheck to generate counterexamples
are shown.  The first column contains the results of using QuickCheck without
shrinking to provide a baseline.  The next two columns show the result of using
QuickCheck with definitions of \ttp{shrink} that limit the lists of fields of
\ttp{B} to 10 (Figure~\ref{lst:newshrink}) and 20 elements
(Figure~\ref{lst:newshrink}, with \ttp{len = 20}), respectively.  (Recall that
with the original definition for shrink from Figure~\ref{fig:initial}, some
counterexamples cause QuickCheck to not return after hours of computation.)  In
each row, we show the maximum, mean, and median number of seconds required by
QuickCheck to generate a counterexample over 100 runs.  In the last row, we show
the mean size of the final value returned, measured by the taking the cumulative
length of the lists of the fields of \ttp{B}.  Notice that while typically
counterexamples are returned in a fraction of a second, there are outliers
taking 21 seconds and 125 seconds for the two tests.  Both tests roughly cut the
size of the counterexample in half; increasing the number of potential
counterexamples can result in a significant performance penalty for this example
without much benefit.

@@

\paragraph{Outline.}
The paper precedes as follows.  We define a type class and some definitions in
Section~\ref{sec:preliminaries} used in the remainder of the paper.  In
Section~\ref{sec:shrinking}, we describe SmartCheck's reduction algorithm for
efficiently shrinking large counterexamples.  In
Section~\ref{sec:generalization}, we turn our attention to the problem of
counterexample generalization.  Here, we present two algorithms for generalizing
counterexamples and constructing formulas for describing them.  We also discuss
a method for property strengthening to discover new counterexamples.  We
describe implementation-specific details in Section~\ref{sec:implementation},
and the results of some experiments in using our tool are presented in
Section~\ref{sec:experiments}.  Related work is given in
Section~\ref{sec:related}.  Finally, concluding remarks and future work are
presented in Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Preliminaries}\label{sec:preliminaries}

To present the algorithms in the following sections, we provide some definitions
here.  Our presentation simplifies the implementation
(Section~\ref{sec:implementation}) somewhat to improve the presentation.
Consider the following types and type class.
%
\begin{code}
type Idx    = Int
type Size   = Int
data SubVal = forall a. SubTypes a => SubVal a

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has five methods.  The \ttp{size} method returns
the size of a value, \ttp{index} returns a sub-value at an breadth-first index
in a value, \ttp{replace} replaces a sub-value at a particular focus,
\ttp{constr} returns the constructor tagging the value, and \ttp{constrs}
returns a list of unique identifiers for the constructors associated with each
variant of the type.

@@

constr  tree = ["B"]
constrs tree = ["B", "L"]
constrs L    = ["B", "L"]
\end{code}
%
\noindent
We will sometimes refer to the unwrapped value returned by \ttp{index a i} as
the \emph{$i$th sub-value of $a$}, so for example, \ttp{B\sub{4} L\sub{6}
  L\sub{8}} is the 4th sub-value of \ttp{tree}.  An invariant of \ttp{index} is
that for any value \ttp{a}, consider the least $i \geq 0$ such that $$\ttp{index
  a i} = \ttp{Nothing}$$ For all $0 \leq j < i$, $$\ttp{index a j} \neq
\ttp{Nothing}$$ We use this invariant as a termination case in recursive
algorithms over the sub-values of a value.

@@

% ------------------------------------------------------------------------------

\section{Shrinking Data}\label{sec:shrinking}
In this section, we describe how to efficiently shrink data values.  Before we
begin, recall the basic idea behind the \ttp{shrink} method of the
\ttp{Arbitrary} class: generate a list of values, each of which is smaller than
the current counterexample.  Each of the new values generated typically bears no
relationship to the original counterexample other than being smaller.

Consider the problem of finding a small counterexample as a graph search
problem.  The set of vertexes are all possible well-typed values.  There is an
edge from vertex $v$ to $v'$ if $v'$ is smaller than $v$ by some measure.  The
goal is to traverse the graph, starting from a counterexample, to find a vertex
from the set of smallest counterexamples in the graph.

The approach taken by \ttp{shrink} is the most naive graph search algorithm: it
simply generates a random set of potential vertexes to check in the graph.
However, it does not exploit that the original counterexample \ttp{cex} is
already in a neighborhood of other likely counterexamples.  So rather than
choose random vertexes, it may be more efficient to test other values that are
closely related to \ttp{cex}.  By restricting the search to \ttp{cex}'s
neighborhood, a smaller counterexample might be missed, and only a local minimum
is returned.  However, with a large state space, finding a global minimum by
random sampling vertexes in the graph is possible but too inefficient, as we saw
from the example in the introduction.

@@

\section{Counterexample Generalization}\label{sec:generalization}

@@

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
An example best explains the first kind of quantified formulas we generate.
Compare the following formula to the counterexample in Figure~\ref{fig:div-cex}:
%
\begin{code}
\ttp{forall x, y . Div (Div x (Div (C 2) (C 13))) y}
\end{code}
%
\noindent
Here, the formula claims that for all well-typed values \ttp{x} and \ttp{y}, the
property fails.  Not only does the generalization assist the programmer's
insight, but it reduces the sheer size of the counterexample, and it can do so
significantly.  We call the kind of formula just shown \emph{universal sub-value
  generalization} and is implemented in SmartCheck.

@@

extrapolate :: SubTypes a
            => a -> (a -> Property) -> IO [Idx]
extrapolate cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs
    | subTrees cex idx idxs
    = extrapolate' (idx+1) idxs
    | otherwise
    = case index cex idx of
        Nothing -> return idxs
        Just v  -> do
          vs <- newVals v scMaxSize scMaxExtrap
          extrapolate' (idx+1)
            (if allFail cex idx vs prop
               then idx:idxs else idxs)

@@

sumTest :: SubTypes a
          => a -> (a -> Property) -> [Idx] -> IO [Idx]
sumTest cex prop exIdxs = sumTest' 1 []
  where
  sumTest' idx idxs
    | subTrees cex idx (exIdxs ++ idxs)
    = sumTest' (idx+1) idxs
    | otherwise
    = case index cex idx of
        Nothing -> return idxs
        Just v  -> do
          vs <- newVals v scMaxSize scConstrMax
          sumTest' (idx+1)
            (if constrFail cex idx vs prop
                  (subConstr v) (subConstrs v)
               then idx:idxs else idxs)

constrFail :: SubTypes a => a -> Idx -> [SubVal]
  -> (a -> Property) -> String -> [String] -> Bool
constrFail cex idx vs prop con allCons =
  constrFail' [con] vs
  where
  constrFail' cons vs'
    | length cons == length allCons
    = True
    | null vs'
    = False
    | go v == Just True
    = constrFail' (c:cons) (tail vs')
    | otherwise
    = constrFail' cons (tail vs')

    where
    v  = head vs'
    c  = subConstr v
    go = fail prop' . replace cex idx
    prop' a = c `notElem` cons ==> prop a
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:constrs}
\end{figure}

The function \ttp{sumTest} takes a counterexample, a property, and a list of
indexes that have already been generalized---perhaps by the extrapolation
algorithm in Figure~\ref{fig:universal}.  The list of course may be empty if no
sub-values have been previously extrapolated.  Like with the other algorithms,
this algorithm also performs a breadth-first search over a counterexample.  In a
call to \ttp{subTrees}, discussed in Section~\ref{sec:universal}, the guard
prevents constructor generalization if the current index is a sub-value of a
previously generalized value.  Otherwise, a list of well-typed new values are
generated by a call to \ttp{newVals}, as shown in Figure~\ref{fig:reduction}.
In the arguments to \ttp{newVals}, we bound the size of values generated with
\ttp{scMaxSize} as before, and bound the number of values generated with the
constant \ttp{scConstrMax}.  Because values are randomly generated, for ``wide''
sum-types (i.e., with a large number of constructors), \ttp{scConstrMax} should
be large enough to ensure with high probability that each variant is generated.

The function \ttp{constrFail} returns true if we replace the sub-value at index
\ttp{idx} in counterexample \ttp{cex} with every possible variant given the type
and construct a counterexample to the property.  There are three guards to the
recursive function \ttp{constrFail'}: the first guard holds if the list of
constructors tagging variants in which a counterexample is found is equal in
size to the list of all possible constructors for the type.  The second guard
tests whether the set of test values is null; if so (and if the first guard
fails), then we have exhausted test values before finding all possible failing
variants.  Third, for a specific sub-value \ttp{v}, we test whether it fails the
property.  If so, we add its constructor to the list of constructors.  Note in
the definition of \ttp{prop'}, we add an additional precondition that the
current constructor is not an element of constructors already seen.  Thus,
\ttp{go v} returns \ttp{Just True} if \ttp{replace cex idx v} passes this
precondition (and any other preconditions of \ttp{prop}), but fails the
property.

@@

\section{Experiments}\label{sec:experiments}

@@

%-------------------------------------------------------------------------------
\section{Related Work}\label{sec:related}

@@

\todo{avoid complexities
  \url{http://stackoverflow.com/questions/14006005/idiomatic-way-to-shrink-a-record-in-quickcheck}}
\cite{}

%-------------------------------------------------------------------------------
\section{Conclusions}\label{sec:conclusions}
